\textbf{Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data}

### 2019-08-19 submission/manuscript.Rmd
changes to interp results

---
title: '**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**'
csl: mbio.csl
fontsize: 11pt
output:
  word_document:
geometry: margin=1.0in
bibliography: references.bib
---

@@

To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC. If the difference between the cross-validation and testing AUROCs was large, then that would indicate that the models were overfit. The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]; however, these differences are relatively small and would not indicate a problem with overfitting.  Subsample experiment results to show what overfitting looks like goes here…

@@

@@

We interpreted the feature importance in the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median absolute feature weights for each OTU in the models [Figure 3]. The OTUs that had the largest median rank and drove the detection of SRNs across the linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 1239, OTU 742), Roseburia (OTU 822, OTU00768), Clostridium (OTU 609), Blautia (OTU 659, OTU 12), Faecalibacterium (OTU 15), Gamella (OTU 426) and Peptostreptococcus (OTU 367) [Figure 3]. Among the five highest ranked OTUs across these three models, it was encouraging that OTU 1239 was shared among each of the models and that 4 OTUs were shared among the five highest ranked OTUs for both of the L2-regularized models.

We interpreted the feature importance in the non-linear models using permutation importance. Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups. We then calculated how much the predictive performance of the model (i.e testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest. The OTUs that had the largest impact and drove the detection of SRNs across the non-linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 50), Ruminococcaceae (OTU 958, 8, 19), Flavonifractor (OTU 629), Howardella (OTU 299), Peptostrepococcus (OTU 367), Clostridium (OTU 58), Anaerostipes (OTU 10), Porphyromonas (OTU 110), Coprobacillus (OTU 477), Parvimonas (OTU 361), Roseburia (OTU 822) [Figure 4]. Among the twenty OTUs with the largest impact across these four models, there was only one OTU (OTU 822) that was shared among all of the models; however, among the tree-based models, we found three OTUs (OTU 367, OTU 58 and OTU 110) to be consistently important. Similarly, in the random forest and XGBoost models, four of the twenty most important OTUs were shared between the models.

When we analyzed the L1-regularized SVM with linear kernel model using permutation importance, the approach picked out many of the same OTUs (e.g. OTU00822, OTU01239, OTU00609) that were important when we interpreted the model using feature rankings based on weights [Figure 3]. Similarly, L2-regularized SVM and L2-regularized logistic regression  each identified many OTUs that were selected by both interpretation methods.

>>>>

---
title: '**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**'
csl: mbio.csl
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
bibliography: references.bib
---

@@

To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC. If the difference between the cross-validation and testing AUROCs was large, then that would indicate that the models were overfit. The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]; however, these differences are relatively small and would not indicate a problem with overfitting.

@@

To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our sample size from 490 to 245, 120, 60, 30 and 15 [Figure S4]. The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when full sample size was used to train and validate the models. These results showed that because microbiome data has many features (6920 OTUs), it is important to train models using appropriate sample sizes to avoid generalizability issues.

@@

We interpreted the feature importance in the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median rank of absolute feature weights of each OTU [Figure 3]. The OTUs that had the largest median rank and drove the detection of SRNs across the linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 1239, OTU 742), Roseburia (OTU 822, OTU00768), Clostridium (OTU 609), Blautia (OTU 659, OTU 12), Faecalibacterium (OTU 15), Gamella (OTU 426) Peptostreptococcus (OTU 367) and others [Figure 3]. Among the highest ranked OTUs across these three models, it was encouraging that many OTUs (e.g. OTU 822, OTU 1239, OTU 426, OTU 609, OTU 50) were shared among all the models.

We interpreted the feature importance in the non-linear models using permutation importance. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups. We then calculated how much the predictive performance of the model (i.e testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest. The OTUs that had the largest impact and drove the detection of SRNs across the non-linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 50), Ruminococcaceae (OTU 958, 8, 19), Flavonifractor (OTU 629), Howardella (OTU 299), Peptostrepococcus (OTU 367), Clostridium (OTU 58), Anaerostipes (OTU 10), Porphyromonas (OTU 110), Coprobacillus (OTU 477), Parvimonas (OTU 361), Roseburia (OTU 822) and others [Figure 4]. Among the twenty OTUs with the largest impact across these four models, there was only one OTU (OTU 822) that was shared among all of the models; however, among the tree-based models, we found three OTUs (OTU 367, OTU 58 and OTU 110) to be consistently important. Similarly, in the random forest and XGBoost models, four of the twenty most important OTUs (OTU 361, 2, 477, 12) were shared between the models.

To highlight the differences between the two interpretation methods, we used permutation importance to interpret linear models [Figure S3]. When we analyzed the L1-regularized SVM with linear kernel model using permutation importance, the approach picked out 17 out of 20 top OTUs (e.g. OTU 822, OTU 1239, OTU 609) that were deemed important when we interpreted the model using feature rankings based on weights [Figure 3]. Similarly, L2-regularized SVM and L2-regularized logistic regression each identified many OTUs that were selected by both interpretation methods, 9 out of 20 and 12 out of 20 respectively.

### 2019-08-19 submission/manuscript.tex
changes to interp results

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

To evaluate the generalizability of each model, we compared the median
cross-validation AUROC to the median testing AUROC. If the difference
between the cross-validation and testing AUROCs is large, then that
would indicate that the models were overfit. difference in median AUROCs
was 0.021 in L1-regularized SVM with linear kernel, followed by SVM with
radial basis function kernel and decision tree with a difference of
0.007 and 0.006, respectively {[}Figure 2{]}; however, these differences
are relatively small and would not indicate a problem with overfitting.

To evaluate the risk for over-optimism of each model, we calculated the
range for each model using 100 splits. The range among the testing AUROC
values within each model varied by 0.23 on average across the seven
models. If we had only done a single split, then there is a risk that we
could gotten lucky or unlucky with the performance of the model. For
instance, the lowest AUROC value of the random forest model was 0.59
whereas the highest was 0.81. These results showed that depending on the
data-split, the testing AUROC values showed great variability {[}Figure
2{]}. Therefore, it is important to employ the hierarchical data splits
that were included in our pipeline to minimize the risk of
over-optimism.

\textbf{Interpretation of each ML model.} Interpretability is the degree
to which humans can understand the reasons behind a model prediction
(31). Because we often use ML models not just to predict a health
outcome but also to learn the ecology behind a disease, model
interpretation becomes crucial for microbiome studies. ML models
decrease in interpretability as they increase in complexity. In this
study we used two methods to interpret models with varying complexity.

We interpreted linear models (L1 and L2-regularized SVM with linear
kernel and L2-regularized logistic regression) using the absolute
feature weights of the trained models. We ranked the absolute values of
the weights of all the OTUs for each data-split {[}Figure 3{]}. We
calculated the median ranks of these features over the 100 data-splits.
In the three linear models, the OTUs that had the largest median rank
and drove the detection of SRNs contained sequences that were most
similar to members of the Lachnospiraceae (OTU 1239, OTU 742), Roseburia
(OTU00822, OTU00768), Clostridium (OTU00609), Blautia (OTU00659,
OTU00012), Faecalibacterium (OTU00015), Gamella (OTU00426) and
Peptostreptococcus (OTU00367) {[}Figure 3{]}. Among the five highest
ranked OTUs across these three models, it was encouraging that OTU 1239
was shared among each of the models and that \ldots{} were shared among
both of the L2-regularized models.

We explained the feature importances in non-linear models; SVM with
radial basis kernel, decision tree, random forest and XGBoost, using a
method called permutation importance on the held-out test data.
Permutation importance analysis is a posthoc explanation of the model
where we randomly permute non-correlated features individually and
groups of highly correlated features together. We then calculate how
much the predictive performance of the model (i.e AUROC values) decrease
when each OTU or group of OTUs is permuted randomly. We ranked the OTUs
based on how much they decreased the median testing AUROC; the OTU with
the largest decrease ranking highest. The top 5 OTUs with the largest
negative impact on testing AUROC overlapped in tree-based models
{[}Figure 4{]}. Specifically, permuting \emph{Peptostreptococcus}
(OTU00367) abundances randomly, dropped the predictive performances the
most in all tree-based methods {[}Figure 4{]}. Decision tree, random
forest and XGBoost models' predictive performance dropped from 0.6
median AUROC to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
respectively {[}Figure 4{]}.

To highlight the differences between the two interpretation methods, we
used permutation importance to interpret linear models as well {[}Figure
S3{]}. L1-regularized SVM with linear kernel picked out some of the same
OTUs (OTU00822, OTU01239, OTU00609) as important in feature rankings
based on weights {[}Figure 3{]} and permutation importance {[}Figure
S3{]}. Similarly, L2-regularized SVM and L2-regularized logistic
regression picked out some of the same OTUs in both interpretation
methods, OTU00659 and OTU00012, respectively. However, for all the
linear models, the rankings of these features were different due to the
collinearity in microbial communities. Collinearity in a microbial
dataset is when one OTU is dependant on another OTU. The feature weights
of correlated OTUs are influenced by one another which makes it
difficult to interpret models using feature weights. Our interpretation
of L2-regularized logistic regression based on feature weight rankings,
showed that an OTU that belongs to \emph{Lachnospiraceae} (OTU00056) was
not among the most important OTUs in making a prediction {[}Figure
3C{]}. However, it was picked out as important when the same model was
interpreted using permutation importance {[}Figure S3C{]}. This is due
to collinearity in the dataset, where \emph{Lachnospiraceae} (OTU00056)
is significantly correlated with \emph{Feacalibacterium} (OTU00015)
which has the highest ranked feature weights {[}Figure 3C{]}. We need to
investigate the relationships among features to identify the true
underlying factors when making a prediction.

@@

We compared the training times of the seven ML models. As the complexity
of a ML model and the number of tuned hyperparameter settings increased
{[}Figures S1-S2{]}, its training times increased as well {[}Figure
5{]}. Linear models trained faster than non-linear models. L1 and
L2-regularized SVM with linear kernel and L2-regularized logistic
regression had the shortest training times with 0.2 hours, (std ± 0.03),
0.2 hours, (std ± 0.02), and 0.2 hours, (std ± 0.02), respectively.
Whereas, a decision tree, SVM with radial basis function kernel, random
forest and XGBoost had training times of 4.4 hours, (std ± 0.3), 59.6
hours, (std ± 8.8), 83.2 hours, (std ± 11.3) and 155.1 hours, (std ± 1),
respectively {[}Figure 5{]}.

@@

@@

Even with interpretable models such as L2-regularized logistic
regression, there are potential pitfalls when it comes to identifying
biomarkers of a disease. As domain experts, we know that
human-associated microbial communities have complex correlation
structures that creates collinearity in the dataset. Collinearity is a
severe problem and needs to be addressed to reliably interpret ML models
(32). In this study we used two different methods to interpret our
linear models; ranking each OTU by (1) their absolute weights in the
trained models and (2) their impact on the predictive performance based
on permutation importance. We observed differences in the OTU rankings
between the two interpretation methods due to collinearity in the
dataset. To avoid misinterpreting the models, once we identify the most
important microbes, we should check for their relationships with other
microbes as well. These relationships will help us generate new
hypotheses about the ecology of the disease. These hypotheses needs to
be tested with follow-up experiments to identify the true biomarkers of
a disease.

@@

We used a balanced CRC dataset to develop ML models with a binary
classification task. We did not evaluate multicategory classification
methods or regression analyses to predict non-binary outcomes. However,
the principles highlighted throughout this study {[}Table 1{]} apply to
all ML modeling tasks with microbiome data. The models we built were
generalizable despite the high number of features microbiome datasets
usually have. The generalization performance of ML models depends on
sample size. The more complex the model, the more data it will need. Our
dataset had 490 samples, however microbiome studies that have smaller
sample sizes would benefit from using less complex models. Our analysis
was limited to shallow learning methods and did not explore deep
learning methods such as neural networks. Microbiome datasets often
suffer from having high dimensionality but low sample sizes which makes
deep learning models prone to overfitting. There are studies that
address overcoming these challenges in biomedical datasets (11, 33, 34),
however sudies that estalish frameworks with microbiome data are
lacking. This would be an interesting direction for future work in
microbiome studies.

@@

\textbf{Data collection and study population.} The data used for this
analysis are stool bacterial abundances and clinical information of the
patients recruited by Great Lakes-New England Early Detection Research
Network study. These data were obtained from Sze et al (35). The stool
samples were provided by recruited adult participants who were
undergoing scheduled screening or surveillance colonoscopy.
Colonoscopies were performed and fecal samples were collected from
participants in four locations: Toronto (ON, Canada), Boston (MA, USA),
Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic health was
labeled by colonoscopy with adequate preparation and tissue
histopathology of all resected lesions. Patients with an adenoma greater
than 1 cm, more than three adenomas of any size, or an adenoma with
villous histology were classified as advanced adenoma. Study had 172
patients with normal colonoscopies, 198 with adenomas and 120 with
carcinomas. Of the 198 adenomas, 109 were identified as advanced
adenomas. Stool provided by the patients was used for 16S rRNA gene
sequencing to measure bacterial population abundances. The bacterial
abundance data was generated by Sze et al, by processing 16S rRNA
sequences in Mothur (v1.39.3) using the default quality filtering
methods, identifying and removing chimeric sequences using VSEARCH and
assigning to OTUs at 97\% similarity using the OptiClust algorithm
(36--38).

@@

\hypertarget{ref-dormann_collinearity:_2013}{}
32. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
\textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
JRG}, \textbf{Gruber B}, \textbf{Lafourcade B}, \textbf{Leitão PJ},
\textbf{Münkemüller T}, \textbf{McClean C}, \textbf{Osborne PE},
\textbf{Reineking B}, \textbf{Schröder B}, \textbf{Skidmore AK},
\textbf{Zurell D}, \textbf{Lautenbach S}. 2013. Collinearity: A review
of methods to deal with it and a simulation study evaluating their
performance. Ecography \textbf{36}:27--46.
doi:\href{https://doi.org/10.1111/j.1600-0587.2012.07348.x}{10.1111/j.1600-0587.2012.07348.x}.

\hypertarget{ref-kocheturov_massive_2019}{}
33. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
2019. Massive datasets and machine learning for computational
biomedicine: Trends and challenges. Ann Oper Res \textbf{276}:5--34.
doi:\href{https://doi.org/10.1007/s10479-018-2891-2}{10.1007/s10479-018-2891-2}.

\hypertarget{ref-kim_improved_2018}{}
34. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
method for prediction of cancer prognosis by network learning. Genes
\textbf{9}:478.
doi:\href{https://doi.org/10.3390/genes9100478}{10.3390/genes9100478}.

\hypertarget{ref-sze_leveraging_2018}{}
35. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors. mBio \textbf{9}:e00630--18.
doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

To evaluate the generalizability of each model, we compared the median
cross-validation AUROC to the median testing AUROC. If the difference
between the cross-validation and testing AUROCs was large, then that
would indicate that the models were overfit. The difference in median
AUROCs was 0.021 in L1-regularized SVM with linear kernel, followed by
SVM with radial basis function kernel and decision tree with a
difference of 0.007 and 0.006, respectively {[}Figure 2{]}; however,
these differences are relatively small and would not indicate a problem
with overfitting.

To evaluate the risk for over-optimism of each model, we calculated the
range of AUROC values for each model using 100 splits. The range among
the testing AUROC values within each model varied by 0.23 on average
across the seven models. If we had only done a single split, then there
is a risk that we could gotten lucky or unlucky with the performance of
the model. For instance, the lowest AUROC value of the random forest
model was 0.593 whereas the highest was 0.81. These results showed that
depending on the data-split, the testing AUROC values showed great
variability {[}Figure 2{]}. Therefore, it is important to employ the
hierarchical data splits that were included in our pipeline to minimize
the risk of over-optimism.

To show the effect of sample size on model generalizability, we compared
cross-validation AUROC values of L2-regularized logistic regression and
random forest models when we subsetted our sample size from 490 to 245,
120, 60, 30 and 15 {[}Figure S4{]}. The range among the cross-validation
AUROC values within both models at lower sample sizes were much larger
than when full sample size was used to train and validate the models.
These results showed that because microbiome data has many features
(6920 OTUs), it is important to train models using appropriate sample
sizes to avoid generalizability issues.

\textbf{Interpretation of each ML model.} Interpretability is the degree
to which humans can understand the reasons behind a model prediction
(31). Because we often use ML models not just to predict a health
outcome but also to learn the ecology behind a disease, model
interpretation becomes crucial for microbiome studies. ML models
decrease in interpretability as they increase in complexity. In this
study we used two methods to help interpret our models.

We interpreted the feature importance in the linear models (L1 and
L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights of each
OTU {[}Figure 3{]}. The OTUs that had the largest median rank and drove
the detection of SRNs across the linear models contained sequences that
were most similar to members of the Lachnospiraceae (OTU 1239, OTU 742),
Roseburia (OTU 822, OTU00768), Clostridium (OTU 609), Blautia (OTU 659,
OTU 12), Faecalibacterium (OTU 15), Gamella (OTU 426) Peptostreptococcus
(OTU 367) and others {[}Figure 3{]}. Among the highest ranked OTUs
across these three models, it was encouraging that many OTUs (e.g.~OTU
822, OTU 1239, OTU 426, OTU 609, OTU 50) were shared among all the
models.

We interpreted the feature importance in the non-linear models using
permutation importance. Whereas the absolute feature weights were
determined from the trained models, here we measured importance using
the held-out test data. Permutation importance analysis is a posthoc
explanation of the model where we randomly permuted non-correlated
features individually and groups of perfectly correlated features across
the two groups. We then calculated how much the predictive performance
of the model (i.e testing AUROC values) decreased when each OTU or group
of OTUs was randomly permuted. We ranked the OTUs based on how much the
median testing AUROC decreased when it was permuted; the OTU with the
largest decrease ranked highest. The OTUs that had the largest impact
and drove the detection of SRNs across the non-linear models contained
sequences that were most similar to members of the Lachnospiraceae (OTU
50), Ruminococcaceae (OTU 958, 8, 19), Flavonifractor (OTU 629),
Howardella (OTU 299), Peptostrepococcus (OTU 367), Clostridium (OTU 58),
Anaerostipes (OTU 10), Porphyromonas (OTU 110), Coprobacillus (OTU 477),
Parvimonas (OTU 361), Roseburia (OTU 822) and others {[}Figure 4{]}.
Among the twenty OTUs with the largest impact across these four models,
there was only one OTU (OTU 822) that was shared among all of the
models; however, among the tree-based models, we found three OTUs (OTU
367, OTU 58 and OTU 110) to be consistently important. Similarly, in the
random forest and XGBoost models, four of the twenty most important OTUs
(OTU 361, 2, 477, 12) were shared between the models.

To highlight the differences between the two interpretation methods, we
used permutation importance to interpret linear models {[}Figure S3{]}.
When we analyzed the L1-regularized SVM with linear kernel model using
permutation importance, the approach picked out 17 out of 20 top OTUs
(e.g.~OTU 822, OTU 1239, OTU 609) that were deemed important when we
interpreted the model using feature rankings based on weights {[}Figure
3{]}. Similarly, L2-regularized SVM and L2-regularized logistic
regression each identified many OTUs that were selected by both
interpretation methods, 9 out of 20 and 12 out of 20 respectively.

Interestingly, for the all the models, except the poorly performing
L1-regularized SVM and decision tree, the variation in the impact of
these important OTUs was minimal. This supports the findings of others
that a classification requires numerous OTUs, rather the presence of a
single or a few of the OTUs (32).

@@

We compared the training times of the seven ML models. As expected, the
training times increased as the complexity of a ML model and the number
of tuned hyperparameter settings increased and the linear models trained
faster than non-linear models {[}Figures S1-S2; Figure 5{]}. When we
subsetted the size of the training dataset, we observed a {[}linear,
quadratic, etc{]} relationship between the size of the dataset and the
training time for the random forest model.

@@

Overview • Establish pipeline that can be generalized to any modeling
method • Linear/non-linear not a bfd • Interpretability tradeoff Point 1
• Importance for transparency, picking the most complex model is not
necessarilty the best model • Good / better / best: Table S1 • ``best''
has multiple meanings Point 2 • Our paper is concerned with developing
models to gain greater biological insights into what's going on • Still
need biological validation and further experimentation Point 3 • These
models show great opportunity to use microbiome to make diagnoses • We
talked about classification, but others have used it in microbiome data
to do regression, mulit-class -- pipeline is the same • Others have used
this on shotgun metagenomic data and collinearity is expected to be an
even larger problem since 5000 genes are all correlated with each other
• Deployment is a different question that requires independent datasets
to test on; we currently lack decent sized parallel datasets where the
same type of sequence data and clinical data have been collected • We
also have a problem with OTU-based approaches Conclusion/Future
direction • ML is awesome and there's always new methods; our framework
gives structure to investigators wanting to try these out. We encourage
reseraches to apply multiple models to their data and report the results
• Things like neural nets are attractive and hold great promise, but we
need a ton more data to apply and they are impossible to interpret

@@

Even with interpretable models such as L2-regularized logistic
regression, there are potential pitfalls when it comes to identifying
biomarkers of a disease. As domain experts, we know that
human-associated microbial communities have complex correlation
structures that creates collinearity in the dataset. Collinearity is a
severe problem and needs to be addressed to reliably interpret ML models
(33). In this study we used two different methods to interpret our
linear models; ranking each OTU by (1) their absolute weights in the
trained models and (2) their impact on the predictive performance based
on permutation importance. We observed differences in the OTU rankings
between the two interpretation methods due to collinearity in the
dataset. To avoid misinterpreting the models, once we identify the most
important microbes, we should check for their relationships with other
microbes as well. These relationships will help us generate new
hypotheses about the ecology of the disease. These hypotheses needs to
be tested with follow-up experiments to identify the true biomarkers of
a disease.

@@

We used a balanced CRC dataset to develop ML models with a binary
classification task. We did not evaluate multicategory classification
methods or regression analyses to predict non-binary outcomes. However,
the principles highlighted throughout this study {[}Table 1{]} apply to
all ML modeling tasks with microbiome data. The models we built were
generalizable despite the high number of features microbiome datasets
usually have. The generalization performance of ML models depends on
sample size. The more complex the model, the more data it will need. Our
dataset had 490 samples, however microbiome studies that have smaller
sample sizes would benefit from using less complex models. Our analysis
was limited to shallow learning methods and did not explore deep
learning methods such as neural networks. Microbiome datasets often
suffer from having high dimensionality but low sample sizes which makes
deep learning models prone to overfitting. There are studies that
address overcoming these challenges in biomedical datasets (11, 34, 35),
however sudies that estalish frameworks with microbiome data are
lacking. This would be an interesting direction for future work in
microbiome studies.

@@

\textbf{Data collection and study population.} The data used for this
analysis are stool bacterial abundances and clinical information of the
patients recruited by Great Lakes-New England Early Detection Research
Network study. These data were obtained from Sze et al (32). The stool
samples were provided by recruited adult participants who were
undergoing scheduled screening or surveillance colonoscopy.
Colonoscopies were performed and fecal samples were collected from
participants in four locations: Toronto (ON, Canada), Boston (MA, USA),
Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic health was
labeled by colonoscopy with adequate preparation and tissue
histopathology of all resected lesions. Patients with an adenoma greater
than 1 cm, more than three adenomas of any size, or an adenoma with
villous histology were classified as advanced adenoma. Study had 172
patients with normal colonoscopies, 198 with adenomas and 120 with
carcinomas. Of the 198 adenomas, 109 were identified as advanced
adenomas. Stool provided by the patients was used for 16S rRNA gene
sequencing to measure bacterial population abundances. The bacterial
abundance data was generated by Sze et al, by processing 16S rRNA
sequences in Mothur (v1.39.3) using the default quality filtering
methods, identifying and removing chimeric sequences using VSEARCH and
assigning to OTUs at 97\% similarity using the OptiClust algorithm
(36--38).

@@

\hypertarget{ref-sze_leveraging_2018}{}
32. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors. mBio \textbf{9}:e00630--18.
doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.

\hypertarget{ref-dormann_collinearity:_2013}{}
33. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
\textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
JRG}, \textbf{Gruber B}, \textbf{Lafourcade B}, \textbf{Leitão PJ},
\textbf{Münkemüller T}, \textbf{McClean C}, \textbf{Osborne PE},
\textbf{Reineking B}, \textbf{Schröder B}, \textbf{Skidmore AK},
\textbf{Zurell D}, \textbf{Lautenbach S}. 2013. Collinearity: A review
of methods to deal with it and a simulation study evaluating their
performance. Ecography \textbf{36}:27--46.
doi:\href{https://doi.org/10.1111/j.1600-0587.2012.07348.x}{10.1111/j.1600-0587.2012.07348.x}.

\hypertarget{ref-kocheturov_massive_2019}{}
34. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
2019. Massive datasets and machine learning for computational
biomedicine: Trends and challenges. Ann Oper Res \textbf{276}:5--34.
doi:\href{https://doi.org/10.1007/s10479-018-2891-2}{10.1007/s10479-018-2891-2}.

\hypertarget{ref-kim_improved_2018}{}
35. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
method for prediction of cancer prognosis by network learning. Genes
\textbf{9}:478.
doi:\href{https://doi.org/10.3390/genes9100478}{10.3390/genes9100478}.
